MIRSVM: Multi-Instance Support Vector Machine with Bag Representatives

### 2017-06-14 ArticleRevision/article/article.tex
Edited Introduction

\begin{abstract}
\textit{Multiple-instance learning} (MIL) is a variation of \textit{supervised learning}, where samples are represented by labeled bags, each containing sets of instances. The individual labels of the samples within a bag are unknown, and labels are assigned based on a multi-instance assumption. One of the major complexities associated with this type of learning is the ambiguous relationship between a bag's label and the instances it contains. This paper proposes a novel support vector machine (SVM) multiple-instance formulation and presents an algorithm with a bag-representative selector that trains the SVM based on bag-level information, named MIRSVM. The contribution is able to identify instances that highly impact classification, i.e. bag-representatives, for both positive and negative bags, while finding the optimal class separation hyperplane. Unlike other multi-instance SVM methods, this approach eliminates possible class imbalance issues by allowing both positive and negative bags to have at most one representative, while highlighting the instances that contribute most to the classification model. The experimental study evaluates and compares the performance of this proposal against $10$ state-of-the-art multi-instance methods over $15$ datasets, and the results are validated through non-parametric statistical analysis. The results indicate that bag-based learners outperform the instance-based and wrapper methods, as well as this proposal's overall superior performance against other multi-instance SVM models. 
\end{abstract}

@@

The difference between MIL and traditional learning is the nature of the data samples. In the traditional setting, each sample is represented by a single feature vector that has its own class label. In multi-instance learning, a sample is considered to be a \textit{bag} that contains multiple instances, and is associated with a single label. The individual labels of the samples within a bag, referred to as \textit{instances}, are unknown, and labels are assigned based on a multi-instance assumption, or hypothesis. Introduced by~\citet{Dietterich1997}, the standard MI assumption states that a bag is labeled as positive if and only if it contains at least one positive instance. Recently, other hypotheses and frameworks have been proposed by~\citet{Foulds2010} to encompass a wider range of applications with MI data.

One of the major complexities associated with multi-instance learning is the ambiguity of the relationship between a bag label and instances within the bag~\citep{Alpaydin2015}. This stems from the standard MIL assumption, where the underlying distribution among instances within positive bags is unknown. All that is known is that at least one instance has a positive label within a positive bag. There have been different attempts to overcome this complexity. One approach involves ``flattenning" the MIL datasets, meaning samples contained in positive bags each adopt a positive label, allowing the use of classical supervised learning techniques~\citep{Ray2005}. This approach assumes that positive bags contain a significant number of positive samples, which may not be the case, causing the classifier to mislabel negative instances within the bag, decreasing the power of the MI model.

To overcome this, a different MIL approach was proposed, where subsets of instances are selected from positive bags for classifier training~\citep{Maron1998,Zhang2001}. One drawback of this type of method is that the resulting training datasets become imbalanced towards positive instances by significantly decreasing the number of negative instances. The performance of these methods deteriorates when more instances are selected as subsets than needed~\citep{Carbonneau2016,Faria2017196,Vanwinckelen2016}. Our proposal aims to deal with this drawback by minimizing class imbalance. This is achieved by optimally selecting bag-representatives from both classes using Support Vector Machines.

Support Vector Machines (SVMs) represent a set of supervised, linear and nonlinear, classification and regression methods that have theoretical foundations on Vapnik-Chervonenkis (VC) theory~\citep{Huang2006,Schoelkopf2002}. SVM models are similar to other machine learning algorithms and techniques, but research has shown that they usually outperform them in terms of computational efficiency, scalability, and robustness against outliers~\citep{Kecman2016,Melki2016}, which makes them a useful data mining tool for various real-world applications.

To address the limitations presented by MIL algorithms, this paper proposes a novel support vector machine formulation with a bag-representative selector, called Multiple-Instance Representative Support Vector Machine (MIRSVM). SVMs are known to perform well when data is limited, therefore combining them with a bag-representative selector aims to remedy class imbalance caused by limited positive bags, without assuming their internal distributions. The algorithm selects bag-representatives iteratively according to the standard MI assumption, ensuring known information, such as the negative bag labels, are fully utilized during the training process. The optimal separating hyperplane between bags is then found with respect to the bag-representatives using a Gaussian kernel and a quadratic programming solver. The optimal set of representatives is found when they have stopped changing from one iteration to the next. The algorithm does not assume any distribution of the instances and is not affected by the number of samples within a bag, making it applicable to a variety of contexts. The key contributions of this work include,
\begin{itemize}
\iitem Reformulating the traditional primal L$1$-SVM problem to optimize over bags, rather than instances, ensuring all the information contained within each bag is utilized during training, while defining bag representative selector criteria.

@@

\iitem Devising a unique bag-representative selection method that makes no presumptions about the underlying distributions of the instances within each bag, while maintaining the default MIL assumption. This approach eliminates the issue of class imbalance caused by techniques such as flattening or subsetting positive instances from each bag. The key feature of MIRSVM is its ability to identify instances (support vectors) within positive and negative bags that highly impact the model.
\end{itemize}

This work is organized as follows. First, the notation used throughout the paper is provided, the MIL problem is formalized, and recent multiple-instance learning methods, traditional support vector machines, and state-of-the-art multi-instance SVM methods are reviewed in Section~\ref{sec:background}. The primal L$1$-SVM is formulated with respect to optimizing over bags, the dual formulation is then derived, along with the conditions for optimality, and the proposed algorithm, MIRSVM, is listed and described in Section~\ref{sec:contribution}. Section~\ref{sec:experiments} presents the experimental environment, including results from various metrics, run-time, and non-parametric statistical analysis over $15$ benchmark MI datasets compared with $10$ state-of-the-art algorithms. Finally, Section~\ref{sec:conclusion} presents the conclusions of this contribution.

\section{Background}\label{sec:background}
This section defines the notation that will be used throughout the paper, and reviews related works on multi-instance learning, traditional support vector machines, and multi-instance support vector machines.

\subsection{Notation}\label{subsec:notation}
Let $\mathcal{D}$ be a training dataset of $n$ bags. Let $\bm{Y} \in \mathcal{D}$ be a vector of $n$ labels corresponding to each bag, having a domain of $\bm{Y} \in \set{-1,\,+1}^n$. Let $\bm{X} \in \mathcal{D}$ be a matrix consisting of $d$ input variables and $m$ instances, having a domain of $\bm{X} \in \mathbb{R}^{m \times d}$. Let $\mathcal{B}$ be the set of bags which contain a number of instances, sometimes of different size, and usually non-overlapping, such as $\bm{B}_I = \{\bm x_{1}, \ldots, \bm x_{|\bm B_I|}\}$ for index set $I \subseteq \set{1,\ldots,n}$, where $|\bm B_I|$ is the number of instances in bag $I$. Using the training dataset $\mathcal{D} = \{(\bm B_1,Y_1), \ldots, (\bm B_n,Y_n)\}$, the goal is to learn a model $f : \bm{X} \times Y$, that assigns a label, $Y_I \in \set{-1,\,+1}$, for each input bag $\bm{B}_I$. The model will then be used to predict the labels of new, unlabeled, and unseen input bags. Table~\ref{tab:Notation} provides a summary of the notation used in this paper.
\begin{table}
\caption{Summary of notation used throughout the paper}\label{tab:Notation}
\begin{tabularx}{\textwidth}{l@{\extracolsep{\fill}}l}
\hline\noalign{\smallskip}
Definition & Notation\\ 
\noalign{\smallskip}\hline\noalign{\smallskip}
Number of Bags & $n$ \\
Number of Instances & $m$ \\
Number of Input Attributes & $d$ \\
Set of Bags & $\mathcal{B} = \set{\bm B_1,\, \ldots,\, \bm B_n}$ \\
Bag Index Set & $I \in \ints_+^n$ \\
Input Space & $\bm{X} \in \mathbb{R}^{m \times d}$ \\
Bag Labels & $\bm{Y} \in \{-1,\,1\}^n$ \\
Input Instance $i$ from Bag $I$ & $\bm{x}_i = (x_{1}, \ldots, x_{d}),\, i \in I$ \\
Individual Instance Label $i$ & $y_i \in \set{-1,+1}$ \\
Bag $I$ & $\bm B_I = \set{\bm x_i \st \forall i \in I}$ \\
\noalign{\smallskip}\hline\noalign{\smallskip}
Full Multi-Instance Training Dataset & $\mathcal{D} = \{(\bm B_1,Y_1), \ldots, (\bm B_n,Y_n)\}$ \\
Full Single-Instance Training Dataset & $\mathcal{S} = \{(\bm x_1,y_1), \ldots, (\bm x_m,y_m)\} $\\
\noalign{\smallskip}\hline
\end{tabularx}
\end{table}

>>>>

\begin{abstract}
\textit{Multiple-instance learning} (MIL) is a variation of \textit{supervised learning}, where samples are represented by labeled bags, each containing sets of instances. The individual labels of the samples within a bag are unknown, and labels are assigned based on a multi-instance assumption. One of the major complexities associated with this type of learning is the ambiguous relationship between a bag's label and the instances it contains. This paper proposes a novel support vector machine (SVM) multiple-instance formulation and presents an algorithm with a bag-representative selector that trains the SVM based on bag-level information, named MIRSVM. The contribution is able to identify instances that highly impact classification, i.e. bag-representatives, for both positive and negative bags, while finding the optimal class separation hyperplane. Unlike other multi-instance SVM methods, this approach eliminates possible class imbalance issues by allowing both positive and negative bags to have at most one representative, which constitute as the most contributing instances to the model. The experimental study evaluates and compares the performance of this proposal against $10$ state-of-the-art multi-instance methods over $15$ datasets, and the results are validated through non-parametric statistical analysis. The results indicate that bag-based learners outperform the instance-based and wrapper methods, as well as this proposal's overall superior performance against other multi-instance SVM models. 
\end{abstract}

@@

The difference between MIL and traditional learning is the nature of the data. In the multi-instance setting, a sample is considered a \textit{bag} that contains multiple \textit{instances} and is associated with a single label. The individual instance labels within a bag are unknown and bag labels are assigned based on a multi-instance assumption, or hypothesis. Introduced by~\citet{Dietterich1997}, the standard MI assumption states that a bag is labeled positive if and only if it contains at least one positive instance. More recently, other hypotheses and frameworks have been proposed by~\citet{Foulds2010} to encompass a wider range of applications with MI data.

One of the major complexities associated with MIL is the ambiguity of the relationship between a bag label and the instances within the bag~\citep{Alpaydin2015}. This stems from the standard MI assumption, where the underlying distribution among instances within positive bags is unknown. There have been different attempts to overcome this complexity, such as ``flattenning" the MIL datasets, meaning instances contained in positive bags each adopt a positive label, allowing the use of classical supervised learning techniques~\citep{Ray2005}. This approach assumes that positive bags contain a significant number of positive samples, which may not be the case, causing the classifier to mislabel negative instances within the bag, decreasing the power of the MI model.

To overcome this, a different MIL approach was proposed, where subsets of instances are selected from positive bags for classifier training~\citep{Maron1998,Zhang2001}. One drawback of this type of method is that the resulting training datasets become imbalanced towards positive instances. Model performance further deteriorates when more instances are selected as subsets than needed~\citep{Carbonneau2016,Faria2017196,Vanwinckelen2016}. Our proposal aims to deal with this drawback by minimizing class imbalance. This is achieved by optimally selecting bag-representatives from both classes using support vector machines.

Support vector machines represent a set of supervised, linear and nonlinear, classification and regression methods that have theoretical foundations on Vapnik-Chervonenkis (VC) theory~\citep{Huang2006,Schoelkopf2002}. SVM models are similar to other machine learning algorithms and techniques, but research has shown that they usually outperform them in terms of computational efficiency, scalability, and robustness against outliers~\citep{Kecman2016,Melki2016}, which makes them a useful data mining tool for various real-world applications.

To address the limitations presented by MIL algorithms, this paper proposes a novel support vector machine formulation with a bag-representative selector, called Multiple-Instance Representative Support Vector Machine (MIRSVM). SVMs are known to perform well when data is limited, therefore combining them with a bag-representative selector aims to remedy class imbalance caused by limited positive bag instances, without assuming the bags' internal distributions. The algorithm selects bag-representatives iteratively according to the standard MI assumption, ensuring known information, such as the negative bag labels, are fully utilized during the training process. The optimal separating hyperplane between bags is then found with respect to the bag-representatives using a Gaussian kernel and a quadratic programming solver. The optimal set of representatives is found when they have converged, or stopped changing from one iteration to the next. The algorithm does not assume any distribution of the instances and is not affected by the number of samples within a bag, making it applicable to a variety of contexts. The key contributions of this work include,
\begin{itemize}
\iitem Reformulating the traditional primal L$1$-SVM problem to optimize over bags, rather than instances, ensuring all the information contained within each bag is utilized during training, while defining bag representative selector criteria.

@@

\iitem Devising a unique bag-representative selection method that makes no presumptions about the underlying distributions of the instances within each bag, while maintaining the default MI assumption. This approach eliminates the issue of class imbalance caused by techniques such as flattening or subsetting positive instances from each bag. The key feature of MIRSVM is its ability to identify instances (support vectors) within positive and negative bags that highly impact the model.
\end{itemize}

This work is organized as follows. First, the notation used throughout the paper is provided, the MIL problem is formalized, and recent MIL methods, traditional support vector machines, and state-of-the-art multi-instance SVM methods are reviewed in Section~\ref{sec:background}. The primal L$1$-SVM is formulated with respect to optimizing over bags, the dual formulation is then derived along with conditions for optimality, and the proposed algorithm, MIRSVM, is listed and described in Section~\ref{sec:contribution}. Section~\ref{sec:experiments} presents the experimental environment, including results from various metrics, run-time, and non-parametric statistical analysis over $15$ benchmark MI datasets compared with $10$ state-of-the-art algorithms. Finally, Section~\ref{sec:conclusion} presents the conclusions of this contribution.

\section{Background}\label{sec:background}
This section defines the notation that will be used throughout the paper, and reviews related works on multi-instance learning, traditional SVMs, and multi-instance support vector machines.

\subsection{Notation}\label{subsec:notation}
Let $\mathcal{D}$ be a training dataset of $n$ bags. Let $\bm{Y} \in \mathcal{D}$ be a vector of $n$ labels corresponding to each bag, having a domain of $\bm{Y} \in \set{-1,1}^n$. Let $\bm{X} \in \mathcal{D}$ be a matrix consisting of $d$ input variables and $m$ instances, having a domain of $\bm{X} \in \mathbb{R}^{m \times d}$. Let $\mathcal{B}$ be the set of bags which contain a number of instances, sometimes of different size and usually non-overlapping, such as $\bm{B}_I = \{\bm x_{1}, \ldots, \bm x_{|\bm B_I|}\}$ for index set $I \subseteq \set{1,\ldots,n}$, where $|\bm B_I|$ is the number of instances in bag $I$. Using the training dataset $\mathcal{D} = \{(\bm B_1,Y_1), \ldots, (\bm B_n,Y_n)\}$, the goal is to learn a model $f : \bm{B} \rightarrow Y$, that assigns a label for each input bag. The model will then be used to predict the labels of new and unseen input bags. Table~\ref{tab:Notation} provides a summary of the notation used in this paper.
\begin{table}
\caption{Summary of notation used throughout the paper}\label{tab:Notation}
\begin{tabularx}{\textwidth}{l@{\extracolsep{\fill}}l}
\hline\noalign{\smallskip}
Definition & Notation\\ 
\noalign{\smallskip}\hline\noalign{\smallskip}
Number of Bags & $n$ \\
Number of Instances & $m$ \\
Number of Input Attributes & $d$ \\
Set of Bags & $\mathcal{B} = \set{\bm B_1,\, \ldots,\, \bm B_n}$ \\
Bag Index Set & $I \in \ints_+^n$ \\
Input Space & $\bm{X} \in \mathbb{R}^{m \times d}$ \\
Bag Labels & $\bm{Y} \in \{-1,1\}^n$ \\
Input Instance $i$ from Bag $I$ & $\bm{x}_i = (x_{1}, \ldots, x_{d}),\, i \in I$ \\
Individual Instance Label $i$ & $y_i \in \set{-1,1}$ \\
Bag $I$ & $\bm B_I = \set{\bm x_i \st \forall i \in I}$ \\
\noalign{\smallskip}\hline\noalign{\smallskip}
Full Multi-Instance Training Dataset & $\mathcal{D} = \{(\bm B_1,Y_1), \ldots, (\bm B_n,Y_n)\}$ \\
Full Single-Instance Training Dataset & $\mathcal{S} = \{(\bm x_1,y_1), \ldots, (\bm x_m,y_m)\} $\\
\noalign{\smallskip}\hline
\end{tabularx}
\end{table}
